---
title: "Storm Data Analysis"
output: html_document
---
##Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

In this analysis, we analyse the existing data set to answer two questions. One of them is regarding the impact of these events on public health and the other one regarding the economic consequences of these events.

We subset the existing data set to filter events that have caused fatalities & injuries (first analysis) and property damage & crop damage (second analysis) greater than zero. We then calculate the mean for both the analysis and filter events that have values above the mean so that we gather higher values. We then plot the analysis using barplots.

In our first analysis, we find that Tornadoes have caused maximum injuries and fatalities across the United States.

In our second analysis, we find that Flash Floods, Thunderstorm Wind, Hail and Floods have caused maximum economic damage to the United States.


##Loading data

```{r load_data, cache=TRUE}
storm_data<-read.csv("repdata-data-StormData.csv.bz2")
```

##Data processing

The values in PROPDMGEXP and CROPDMGEXP are substitued with the right numeric values from the vector below.

```{r process_data,cache=TRUE}
multiply <- c("?"= "NA" ,"-"= "NA","+"= "NA","0"=0,"1" =10, "2"=100 ,"3"=1000, "4"=10000, "5"=100000, "6"=1000000, "7"=10000000,"8"=100000000, "H"=100, "h"=100, "K"=1000, "M"=1000000, "m"=1000000,"B"=1000000000 )
storm_data$PROPDMGEXP<-multiply[as.numeric(storm_data$PROPDMGEXP)]
storm_data$CROPDMGEXP<-multiply[as.numeric(storm_data$CROPDMGEXP)]


#Multiply values
storm_data$PROPDMG<-storm_data$PROPDMG * as.numeric(storm_data$PROPDMGEXP)
storm_data$CROPDMG<-storm_data$CROPDMG * as.numeric(storm_data$CROPDMGEXP)
```

##Analysis

###Which event type has caused harm to population health?

We filter data containing INJURIES and FATALITIES and aggregate by the sum to calculate total Fatalities and Injuries.

We then filter events whose INJURIES and FATALITIES are greater than the mean of the same, respectively.

```{r harm_population_health}
library(dplyr)
population_health<-aggregate(cbind(INJURIES,FATALITIES)~EVTYPE,data=storm_data,FUN=sum)

#Filter rows where injuries and fatalities >0
population_sub<-filter(population_health,INJURIES>0 | FATALITIES>0)

#Filter rows whose injuries and fatalities are above the mean to gather max values
population_sub2<-filter(population_sub,(INJURIES>mean(INJURIES) | FATALITIES>mean(FATALITIES)))
```

We now look at the cleaned data set depicting only the EVENT TYPE and the total INJURIES and FATALITIES caused by each event.

```{r display1}
population_sub2
```


###Which types of events have the greatest economic consequences?

We filter data containing PROPDMG and CROPDMG and aggregate by the sum to calculate total damages caused to the country.

We then filter events whose PROPDMG and CROPDMG are greater than the mean of the same, respectively.

```{r economic_damage}
damage_data<-aggregate(cbind(PROPDMG,CROPDMG)~EVTYPE,data=storm_data,FUN=sum)

#Filter rows where damage>0
damage_data_sub<-filter(damage_data,PROPDMG>0 | CROPDMG>0)

#Filter rows whose damage is above the mean to gather max values
max_damage_data<-filter(damage_data_sub,(PROPDMG>mean(PROPDMG) | CROPDMG>mean(CROPDMG)))
```

We now look at the cleaned data set depicting only the EVENT TYPE and the total PROPDMG and CROPDMG caused by each event.

```{r display2}
max_damage_data
```

##Results

We now plot the results of the two questions using barplots.

###Which event type has caused harm to population health?

```{r plot_graph1}
#Plot
par(mar=c(1,15,1,0),cex=0.7)

barplot(population_sub2$INJURIES,col=terrain.colors(nrow(population_sub2)),names.arg = population_sub2$EVTYPE,horiz = TRUE,las=1,main="Total Injuries across the U.S")
legend("topright", 
       legend = c("xlab=Total Injuries", "ylab=Event Type"))

barplot(population_sub2$FATALITIES,col=terrain.colors(nrow(population_sub2)),names.arg = population_sub2$EVTYPE,horiz = TRUE,las=1,main="Total Fatalities across the U.S")
legend("topright", 
       legend = c("xlab=Total Fatalities", "ylab=Event Type"))
```

We can see that Tornadoes have caused maximum injuries and fatalities.

###Which types of events have the greatest economic consequences?

```{r plot_graph2}
#Plot
par(mar=c(1.8,13,1.5,0),mfrow=c(2,1),cex=0.7)

barplot(max_damage_data$PROPDMG,col=terrain.colors(nrow(max_damage_data)),names.arg = max_damage_data$EVTYPE,horiz = TRUE,las=1,main="Total Property Damage across the U.S")
legend("topright", 
       legend = c("xlab=Total Property Damage", "ylab=Event Type"))

barplot(max_damage_data$CROPDMG,col=terrain.colors(nrow(max_damage_data)),names.arg = max_damage_data$EVTYPE,horiz = TRUE,las=1,main="Total Crop Damage across the U.S")
legend("topright", 
       legend = c("xlab=Total Crop Damage", "ylab=Event Type"))
```

We can see that Flash Floods, Thunderstorm Wind, Hail and Floods have caused maximum economic damage to the U.S.

##Conclusion

We find that Tornadoes have caused maximum injuries and fatalities across the United States and Flash Floods, Thunderstorm Wind, Hail and Floods have caused maximum economic damage to the United States.

Thus, resources have to be mobilized efficiently to battle these extreme weather conditions.

